---
title: "Engelmann-spruce-monoterpenes"
author: "Ehsan Khedive"
date: "2024-08-08"
output: word_document
---

    ***Knit the file to receive the supplemental material***

```{r setup, include=FALSE,}
knitr::opts_chunk$set(echo = TRUE)
install.packages(setdiff(c("car" , "ggplot2", "dplyr", "tidyr", "readr", "effects", "multcomp", "GGally", "gridExtra","rcompanion", "ggpattern", "officer", "flextable", "broom"),
                         rownames(installed.packages())))
library(car)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(effects)
library(multcomp)
library(GGally)
library(gridExtra)
library(rcompanion)
library(ggpattern)
library(officer) # write docx file
library(flextable) # make tables in docx
library(broom) # Extract output coefficients from the model
```

    ***Data for the codes under this heading are not available due to the fact that not all the compounds were studied in this paper***

# 1. Data preparation

## 1.1. Reading datasets

```{r, include=FALSE, eval=FALSE}
setwd("...")
files <- list.files(pattern = "*.csv")

data_list <- lapply(files, function(file) {
  # Read the file
  df <- read_csv(file, skip = 2)
  
  # Identify problematic rows where "RT" is not a number
  problematic_indices <- which(is.na(as.numeric(df$RT)))
  
  # Create a vector of indices to remove (problematic rows + row above each)
  indices_to_remove <- unique(c(problematic_indices, problematic_indices - 1))
  
  # Remove the identified rows
  df <- df[-indices_to_remove, ]
  
  return(df)
})
combined_data <- do.call(rbind, data_list)
rm(data_list,files)
```

## 1.2. Data handling

```{r, include=FALSE, eval=FALSE}
# Remove rows with misplaced columns
combined_data$Mass <- as.character(combined_data$Mass)
rows_with_number <- grepl("\\d", combined_data$Mass)
Pooled_filtered <- combined_data[!rows_with_number, ]
rm(rows_with_number)

Pooled_filtered$`CAS ID` <- sub("^0+", "", Pooled_filtered$`CAS ID`) # Remove leading zeros from "CAS.ID" column
Pooled_filtered$File <- sub(".*\\\\", "", Pooled_filtered$File) # Remove address from File field

# Refine Label column
Pooled_filtered$Label <- sub("^[^ ]+ [^ ]+ ", "", Pooled_filtered$Label)
Pooled_filtered$Label <- sub(";.*$", "", Pooled_filtered$Label)

# Refine Score (lib)

Pooled_filtered <- Pooled_filtered %>%
  filter(!is.na(`Score (Lib)`),          # Remove NAs
         !grepl("[^0-9.]", `Score (Lib)`), # Remove rows with any character (non-numeric)
         as.numeric(`Score (Lib)`) >= 60)  # Remove numbers below 60


Pooled_filtered$`CAS ID` <- paste("CAS", Pooled_filtered$`CAS ID`, sep = "") #adding CAS to CAS ID column
Pooled_filtered$File <- gsub("P|\\.D|\\[MCF50\\]", "", Pooled_filtered$File)
Pooled_filtered$File <- gsub("[^0-9]", "", Pooled_filtered$File) # Retain only the numbers in File column

```

## 1.3. Correct wrong sample labelings

```{r, include=FALSE, eval=FALSE}
Pooled_filtered$File <- gsub("5613", "50613", Pooled_filtered$File) # Correct formatting for tree #95 date 0613
Pooled_filtered <- Pooled_filtered[Pooled_filtered$File != "3606132", ] # Remove duplicate for sample 360613
# Almq
Pooled_filtered$File <- gsub("630823", "630828", Pooled_filtered$File) # Correct date
Pooled_filtered$File <- gsub("670827", "670828", Pooled_filtered$File) # Correct date
Pooled_filtered$File <- gsub("730720", "730920", Pooled_filtered$File) # Correct date
# Cameron pass
Pooled_filtered$File <- gsub("105", "06", Pooled_filtered$File) # Correct tree tag
Pooled_filtered$File <- gsub("120610", "080610", Pooled_filtered$File) # Correct tree tag
Pooled_filtered$File <- gsub("070913", "070917", Pooled_filtered$File) # Correct date
# Empire
Pooled_filtered$File <- gsub("830602", "830620", Pooled_filtered$File) # Correct date
Pooled_filtered$File <- gsub("790716", "790710", Pooled_filtered$File) # Correct date
# Grant
Pooled_filtered$File <- gsub("500708", "500703", Pooled_filtered$File) # Correct date
Pooled_filtered$File <- gsub("470710", "470712", Pooled_filtered$File) # Correct date
Pooled_filtered$File <- gsub("580812", "580612", Pooled_filtered$File) # Correct date
Pooled_filtered$File <- gsub("460920", "400921", Pooled_filtered$File) # Correct date
# Rabbit ears
Pooled_filtered$File <- gsub("240819", "240918", Pooled_filtered$File) # Correct date
Pooled_filtered$File <- gsub("290613", "290619", Pooled_filtered$File) # Correct date
Pooled_filtered$File <- gsub("300613", "300619", Pooled_filtered$File) # Correct date
Pooled_filtered$File <- gsub("320613", "320619", Pooled_filtered$File) # Correct date
Pooled_filtered$File <- gsub("330613", "330619", Pooled_filtered$File) # Correct date
Pooled_filtered$File <- gsub("340613", "340619", Pooled_filtered$File) # Correct date
Pooled_filtered$File <- gsub("350613", "350619", Pooled_filtered$File) # Correct date
Pooled_filtered$File <- gsub("360613", "360619", Pooled_filtered$File) # Correct date
Pooled_filtered$File <- gsub("370613", "370619", Pooled_filtered$File) # Correct date
Pooled_filtered$File <- gsub("380613", "380619", Pooled_filtered$File) # Correct date
Pooled_filtered$File <- gsub("380708", "380709", Pooled_filtered$File) # Correct date


Pooled_filtered$File <- gsub("320728", "070728", Pooled_filtered$File) # Correct date

```

## 1.4. Import duplicate samples

```{r, include=FALSE, eval=FALSE}
Duplicates <- read_csv("Duplicates.csv", skip = 2)
# No misplaced columns was detected

Duplicates <- Duplicates[Duplicates$`Show/Hide` == "True", ] # Remove undetected peaks
Duplicates <- Duplicates[!is.na(Duplicates$File), ]

Duplicates$`CAS ID` <- sub("^0+", "", Duplicates$`CAS ID`) # Remove leading zeros from "CAS.ID" column
Duplicates$File <- sub(".*\\\\", "", Duplicates$File) # Remove address from File field

# Refine Label column
Duplicates$Label <- sub("^[^ ]+ [^ ]+ ", "", Duplicates$Label)
Duplicates$Label <- sub(";.*$", "", Duplicates$Label)


Duplicates$`CAS ID` <- paste("CAS", Duplicates$`CAS ID`, sep = "") #adding CAS to CAS ID column
Duplicates$File <- gsub("[^0-9]", "", Duplicates$File) # Retain only the numbers in File column

# Correct labels
{Duplicates$File <- gsub("6908132", "650813", Duplicates$File)
Duplicates$File <- gsub("760920", "670920", Duplicates$File)
Duplicates$File <- gsub("690730", "970730", Duplicates$File)
Duplicates$File <- gsub("080628", "180628", Duplicates$File)
Duplicates$File <- gsub("960620", "940620", Duplicates$File)
Duplicates$File <- gsub("3606132", "310619", Duplicates$File)}

```

## 1.5. Merge dataset with duplicates
The final version of raw data set was exported in this chunk and then manually corrected for the labels.

```{r, include=FALSE, eval=FALSE}
Raw_final <- rbind(subset(Pooled_filtered, select = c(RT, File, Label, `Score (Lib)`, Formula, `CAS ID`, Precursor, Area, `Reference mass`, `Base Peak`, Ions)), subset(Duplicates, select = c(RT, File, Label, `Score (Lib)`, Formula, `CAS ID`, Precursor, Area, `Reference mass`, `Base Peak`, Ions)))

Raw_final$RT <- as.numeric(Raw_final$RT)
Raw_final <- Raw_final %>%
  mutate(File = as.character(File)) %>%
  separate(File, into = c("TreeID", "Date"), sep = "(?<=.)(?=\\d{4}$)", remove = FALSE)

Raw_final <- Raw_final %>%
  mutate(Site = factor(case_when(
    grepl("^(79|[8-9][0-9])", TreeID) ~ "Empire",
    grepl("^(59|[6-7][0-9])", TreeID) ~ "Alma",
    grepl("^(39|[4-5][0-9])", TreeID) ~ "Grant",
    grepl("^(19|[2-3][0-9])", TreeID) & !grepl("200", TreeID) ~ "Rabbit ears",
    grepl("^\\d", File) ~ "Cameron pass",
    TRUE ~ NA_character_
  )))

Treatments <- read.csv("Treatments.csv") # Load treatment codes
Raw_final$TreeID <- as.numeric(Raw_final$TreeID)
Raw_final$Date <- as.numeric(Raw_final$Date)

Raw_final <- merge(Raw_final, Treatments, by = "TreeID", all.x = TRUE)
write.csv(Raw_final, "Raw_final.csv")
```

## 1.6. Import Correctly labeld dataset with grouping and damage status

```{r, include=FALSE, eval=FALSE}
Corrected_labels <- read_csv("Corrected_labels.csv")

# Add compound groups
Groupings <- read_csv("Groups.csv")
Corrected_labels$Group <- Groupings$Group[match(Corrected_labels$Corrected_Label, Groupings$Corrected_Label)]
Corrected_labels$Group[is.na(Corrected_labels$Group)] <- "unknown"

# Add tree status, resinosis and Entry holes
TreeStatus <- read_csv("Tree_status.csv")
Corrected_labels$Status <- TreeStatus$Status[match(Corrected_labels$TreeID, TreeStatus$TreeID)]
Corrected_labels$Status[is.na(Corrected_labels$Status)] <- "Control"
Corrected_labels$Resinosis <- TreeStatus$Resinosis[match(Corrected_labels$TreeID, TreeStatus$TreeID)]
Corrected_labels$Entryholes <- TreeStatus$`Entryholes-sqm`[match(Corrected_labels$TreeID, TreeStatus$TreeID)]

# Add site climate
Climatedata <- read.csv("Site_climate.csv")
Corrected_labels$Precipitation <- Climatedata$Precipitation[match(Corrected_labels$Site, Climatedata$Site)]
Corrected_labels$Mean_temperature <- Climatedata$Mean_temperature[match(Corrected_labels$Site, Climatedata$Site)]
Corrected_labels$Maximum_VPD <- Climatedata$Maximum_VPD[match(Corrected_labels$Site, Climatedata$Site)]
Corrected_labels$Elevation <- Climatedata$Elevation[match(Corrected_labels$Site, Climatedata$Site)]
Corrected_labels <- Corrected_labels %>%
  mutate(Day_of_year = case_when(
    Date %in% c(607:613) ~ "160",
    Date %in% c(618:623) ~ "170",
    Date %in% c(628:703) ~ "180",
    Date %in% c(708:712) ~ "190",
    Date %in% c(718:723) ~ "200",
    Date %in% c(728:801) ~ "210",
    Date %in% c(812:814) ~ "225",
    Date %in% c(827:828) ~ "240",
    Date %in% c(917:921) ~ "260",
    TRUE ~ NA_character_ # Default for dates not matching any condition
  ))
Corrected_labels$Day_of_year <- as.numeric(Corrected_labels$Day_of_year)
Treatments <- read.csv("Treatments.csv") # Load treatment codes
Corrected_labels <- merge(Corrected_labels, Treatments[, c("TreeID", "Diameter")], by = "TreeID", all.x = TRUE)

# Aggregate multiple peaks in one cell
Corrected_labels_sumed <- aggregate(Area ~ Corrected_Label + TreeID + Day_of_year, 
                                     data = Corrected_labels, 
                                     FUN = sum)
unique_rows <- (Corrected_labels[!duplicated(Corrected_labels[, c("Corrected_Label", "TreeID", "Day_of_year")]),setdiff(names(Corrected_labels), "Area") ])
merged_data <- merge(Corrected_labels_sumed, unique_rows, 
                     by = c("Corrected_Label", "TreeID", "Day_of_year"))

rm(Treatments,TreeStatus, Groupings,unique_rows,Corrected_labels, Corrected_labels_sumed)
```

## 1.7. Import sample weight dataset and add to the main dataset

```{r, include=FALSE, eval=FALSE}
Sampleweight <- read.csv("Sample Weights.csv")
Sampleweight <- Sampleweight %>%
  mutate(Day_of_year = case_when(
    Date %in% c(607:613) ~ "160",
    Date %in% c(618:623) ~ "170",
    Date %in% c(628:703) ~ "180",
    Date %in% c(708:712) ~ "190",
    Date %in% c(718:723) ~ "200",
    Date %in% c(728:801) ~ "210",
    Date %in% c(812:814) ~ "225",
    Date %in% c(827:828) ~ "240",
    Date %in% c(917:921) ~ "260",
    TRUE ~ NA_character_ # Default for dates not matching any condition
  ))
Sampleweight$Day_of_year <- as.numeric(Sampleweight$Day_of_year)
Sampleweight <- Sampleweight[Sampleweight$Comment != "Wrong derivatization", ]

merged_data <- merged_data %>%
  left_join(Sampleweight %>% dplyr::select(TreeID, Day_of_year, Phloem..mg., Solvent..ul.), 
            by = c("TreeID", "Day_of_year"))


## Standard solvent concentration
merged_data$Value <- (merged_data$Area*merged_data$Solvent..ul.)/200

## Add estimate sample weights to samples without phloem weight. I weighed samples in a range of 100 mg to 120 mg with average of 110 mg
merged_data$Phloem..mg.[is.na(merged_data$Phloem..mg.)] <- 110

rm(Sampleweight)
```

## 1.8. Extract monoterpen compounds from the main dataset for this manuscript and write them to a file
```{r, include=FALSE, eval=FALSE}

Monoterpenes <- merged_data[!is.na(merged_data$Group) & 
                                 merged_data$Group %in% c("Monoterpene"), ]

# Remove unused columns
Monoterpenes <- Monoterpenes %>%
  dplyr::select(-c(RT, File, Date, Group, Area))

# Replace outliers with 99% percentile using Winzorization
Monoterpenes <- Monoterpenes %>%
  group_by(Corrected_Label) %>%
  mutate(
    upper_quantile = quantile(Value, 0.99, na.rm = TRUE),
     Value_Winz = ifelse(Value > upper_quantile, upper_quantile, Value)
  ) %>%
  ungroup()

# Clean data from missing samples
Monoterpenes <- Monoterpenes[!is.na(Monoterpenes$Solvent..ul.), ]

write.csv(Monoterpenes,"Monoterpenes.csv")

```



                                                                ***Readers/Reviewers can run the code chunks from this point forward.***

# 2. Read and prepared monoterpenes dataset for analysis

## 2.1. Read monoterpenes data file
```{r Import data}
setwd("...")
Monoterpenes <- read_csv("Monoterpenes.csv")[,-1]
Climatedata <- read.csv("Site_climate.csv")
```


## 2.2. Response curve calculation and quantitiation
Concentrations are described as nanoliters per milliliters of solvent, but then converted to microgram per gram of tissue based on the solvent volume and monoterpene density.

```{r Quantify concentrations}
Standards <- read.csv("Standards.csv", header = TRUE)

# Create a list to store the models for each monoterpene
standard_list <- list()

# Loop over each monoterpene to fit the two calibration models
for(compound in unique(Standards$Monoterpene)){
  
  # Subset data for the current compound
  compound_data <- Standards %>% filter(Monoterpene == compound)
  
  # Define the two subsets based on concentration calibration points (1 to 250 nanoliters per milliliter):
  low_cal_data  <- compound_data %>% filter(Concentration %in% c(1, 5, 25))
  high_cal_data <- compound_data %>% filter(Concentration %in% c(25, 50, 250))
  
  # Fit the linear regression models (using Area as predictor)
  model_low  <- lm(Concentration ~ Area, data = low_cal_data)
  model_high <- lm(Concentration ~ Area, data = high_cal_data)
  
  # Save the models into the list (using the compound name as key)
  standard_list[[compound]] <- list(low = model_low, high = model_high)
}

# Add surrogate compounds
standard_list$`Linalool oxide` <- standard_list$Linalool
standard_list$`.alpha.-Phellandrene` <- standard_list$`.gamma.-Terpinene`
standard_list$`.alpha.-Terpinene` <- standard_list$`.gamma.-Terpinene`
standard_list$`.beta.-Ocimene` <- standard_list$`.beta.-Myrcene`
standard_list$`.beta.-Phellandrene` <- standard_list$Limonene
standard_list$`3-Thujene` <- standard_list$`3-Carene`
standard_list$Sabinene <- standard_list$`3-Carene`

# Define a function that predicts concentration given a compound and a peak area. If the peak area falls below 2,500,000 the concentration will be calculated with the low model and if it's above 2,500,000 it will be calculated with the high model.
predict_concentration <- function(compound, area, threshold = 2500000) {
  # Retrieve the two models for the given compound
  models <- standard_list[[compound]]
  if(is.null(models)){
    stop("Compound not found in the calibration Standards")
  }
  
  # Choose the model based on the area threshold
  if(area < threshold) {
    pred <- predict(models$low, newdata = data.frame(Area = area))
  } else {
    pred <- predict(models$high, newdata = data.frame(Area = area))
  }
  return(pred)
}

# Calculate concentrations based on the response curve and winzorized peak area
Monoterpenes$Concen <- mapply(predict_concentration, 
                                           compound = Monoterpenes$Corrected_Label, 
                                           area = Monoterpenes$Value_Winz)

# Calculate weight based on compound density

Monoterpene_density <- data.frame(Monoterpene = c(".alpha.-Phellandrene", ".alpha.-Pinene", ".alpha.-Terpinene", 
".beta.-Myrcene", ".beta.-Ocimene", ".beta.-Phellandrene", ".beta.-Pinene", 
".gamma.-Terpinene", "3-Thujene", "Camphene", "Limonene", "Linalool oxide", 
"Linalool", "p-Cymene", "Sabinene", "Terpinolene"), Density = c(0.846, 0.858, 0.841, 
  0.794, 0.789, 0.846, 0.872, 
  0.841, 0.86, 0.842, 0.841, 0.95, 
  0.858, 0.857, 0.846, 0.864))

# Calculate tissue concentration
Monoterpenes$ConcentrationUGpGFW <- (Monoterpenes$Concen * 0.2 * 
                            Monoterpene_density$Density[match(Monoterpenes$Corrected_Label, 
                                     Monoterpene_density$Monoterpene)])/Monoterpenes$Phloem..mg.*1000
rm(list = setdiff(ls(),c("Monoterpenes","Climatedata")))

```

```{r Pivot monoterpene dataset}
# Change data set to wide format
Monoterpenes_Wide <- Monoterpenes %>%
                dplyr::select(-c(Solvent..ul.,Value, upper_quantile, Value_Winz, Concen))
# Define factor variables
Monoterpenes_Wide$Status <- as.factor(Monoterpenes_Wide$Status)
Monoterpenes_Wide$Day_of_year <- as.factor(Monoterpenes_Wide$Day_of_year)
Monoterpenes_Wide$Site <- as.factor(Monoterpenes_Wide$Site)
Monoterpenes_Wide$Location <- as.factor(Monoterpenes_Wide$Location)
Monoterpenes_Wide$Treatment <- as.factor(Monoterpenes_Wide$Treatment)

# Pivot the dataset for monoterpenes
Monoterpenes_Wide <- Monoterpenes_Wide %>%
  pivot_wider(
    names_from = Corrected_Label, # Use unique values from the Corrected_Label column as new column names
    values_from = ConcentrationUGpGFW,          # Use the 'value' column for the data in the new columns
    values_fn = sum,              # Function to sum duplicate values
    values_fill = 0
  )

# Calculate total concentration of monoterpenes
Monoterpenes_Wide$Total <- rowSums(Monoterpenes_Wide[15:30],)

```

## Plot output for individual monoterpenes and correlation

```{r Individual monoterpene analysis}
monoterpene_vars <- c(".alpha.-Phellandrene", ".alpha.-Pinene", ".alpha.-Terpinene", 
".beta.-Myrcene", ".beta.-Ocimene", ".beta.-Phellandrene", ".beta.-Pinene", 
".gamma.-Terpinene", "3-Thujene", "Camphene", "Limonene", "Linalool oxide", 
"Linalool", "p-Cymene", "Sabinene", "Terpinolene")

# Initialize an empty list to store predicted data frames for each compound
predicted_list <- list()

# Loop over each monoterpene variable
for (var in monoterpene_vars) {
  # Create a formula dynamically; note the use of backticks to account for unusual characters in variable names.
  formula <- as.formula(paste0("`", var, "` ~ Day_of_year * Status"))
  
  # Fit the model (using a gaussian family here, which is equivalent to lm)
  model <- glm(formula, data = Monoterpenes_Wide, family = gaussian())
  
  # Get predicted values from the model
  Monoterpenes_Wide$Predicted <- predict(model)
  
  # Create a temporary dataframe containing the key columns and the current compound name.
  temp_df <- Monoterpenes_Wide %>%
    dplyr::select(Day_of_year, Status, Predicted) %>%
    mutate(Compound = var)
  
  # Store the temporary dataframe in the list using the compound name as key.
  predicted_list[[var]] <- temp_df
}

# Combine predictions for all compounds into one data frame
predicted_all <- bind_rows(predicted_list)

# Figure S1. Concentration of individual monoterpenes in phloem samples during the growth season
Figure_S1 <- ggplot(predicted_all, aes(x = Day_of_year, y = Predicted, color = Status, group = Status)) +
  geom_line(linewidth = 1) +
  geom_point() +
  facet_wrap(~ Compound, scales = "free_y") +
  labs(title = "Figure S1. Individual Monoterpene Concentrations During the Growth Season",
       x = "Day of Year",
       y = expression(Concentration ~ (mu*g/g ~ FW)),
       color = "Status") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

# Save the plot as a PDF with custom dimensions (e.g., width = 16 inches, height = 12 inches)
ggsave("Figure_S1.pdf", plot = Figure_S1, width = 16, height = 12, units = "in")

# Figure S2. Correlation among individual monoterpenes in phloem samples during the growth season
Figure_S2 <- ggpairs(
  Monoterpenes_Wide[ , 15:30],
  lower = list(continuous = wrap("points", alpha = 0.5, size = 1)),
  upper = list(continuous = wrap("cor", size = 4, method = "pearson")),
  diag = list(continuous = wrap("densityDiag")),
  progress = FALSE
)+
  ggtitle("Figure S2. Pairwise correlations among monoterpenes") +
  theme(plot.title = element_text(hjust = 0.5))

ggsave("Figure_S2.pdf", plot = Figure_S2, width = 16, height = 12, units = "in")

rm(list = setdiff(ls(), c("Climatedata", "Monoterpenes", "Monoterpenes_Wide")))
```


    ***The rest of the code chunks are arranged based on the results section***

# 3.1. Constitutive and induced defense as a function of climate factors


## Site effect
```{r Site effect}
Control_trees <- Monoterpenes_Wide[Monoterpenes_Wide$Treatment == "Control",]
Control_trees <- Control_trees[,c(1:6,9:13,31)]
model_Site <- lm(log(Total) ~ Day_of_year + Site, 
                       data = Control_trees)

# Table S1. Results of complete block ANOVA of site effect on log of total monoterpene concentration
coef_df <- tidy(model_Site)
fit_df <- glance(model_Site)

coef_df <- coef_df[, c("term", "estimate", "std.error", "statistic", "p.value")]
names(coef_df) <- c("Term", "Estimate", "SE", "t", "p-value")
coef_df$`p-value` <- format.pval(
  coef_df$`p-value`,
  digits = 3,
  eps = 0.001
)
ft_coef <- flextable(coef_df)
ft_coef <- autofit(ft_coef)


fit_out <- data.frame(
  Statistic = c(
    "Residual SE",
    "R²",
    "Adjusted R²",
    "F statistic",
    "Model p-value"
  ),
  Value = c(
    round(fit_df$sigma, 3),
    round(fit_df$r.squared, 3),
    round(fit_df$adj.r.squared, 3),
    round(fit_df$statistic, 2),
    format.pval(fit_df$p.value, eps = 0.001)
  )
)

ft_fit <- autofit(flextable(fit_out))

doc <- read_docx()

doc <- body_add_par(
  doc,
  "Table S1. Results of complete block ANOVA of site effect on log of total monoterpene concentration",
  style = "Normal"
)
doc <- body_add_par(doc, "Model: log(Total monoterpenes) ~ Day of year + Site", style = "Normal")
doc <- body_add_par(doc, "Model coefficients", style = "centered")
doc <- body_add_flextable(doc, ft_coef)
doc <- body_add_par(doc, "Model fit statistics", style = "centered")
doc <- body_add_flextable(doc, ft_fit)

# Extract the effect of Site from the model
Site_effect <- as.data.frame(effect("Site", model_Site))

# Tukey's HSD Test
tukey_res <- glht(model_Site, linfct = mcp(Site = "Tukey"))
# Add letters to Site_effect dataframe
Site_effect$Significance <- cld(tukey_res)$mcletters$Letters

# Create a ggplot for the Site effect
Figure_2 <- ggplot(Site_effect, aes(x = Site, y = exp(fit))) +
  geom_point(aes(color = Climatedata$Mean_temperature, size = Climatedata$Elevation),) +  # Color based on Temperature
  scale_size_continuous(range = c(5, 12)) + # Adjust min and max sizes
  scale_color_gradient(low = "blue", high = "red") +
  geom_errorbar(aes(ymin = exp(lower), ymax = exp(upper)), width = .10, color = "darkblue") +  # Confidence intervals
  geom_text(aes(label = Significance, y = exp(upper)-10), position = position_jitter(width = 0.0, height = 0.0), size = 10, color = "black") +  # Add significance letters
  labs(
    x = "Site",
    y = expression('Monoterpenes Concentration' ~ (mu*g/g ~ FW)),
    color = "Temperature (\u00B0C)",
    size = "Elevation (M)"
  ) +
  theme_minimal() +
  theme(
    text = element_text(family = "Helvetica"),
    axis.text.x = element_text(size = 20, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 20, hjust = 1),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 18),
    legend.title = element_text(size = 20),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
  )

ggsave("Figure_2.pdf", plot = Figure_2, width = 30, height = 20, units = "cm")

rm(list = setdiff(ls(), c("Climatedata", "Monoterpenes", "Monoterpenes_Wide", "Control_trees", "doc")))
```

## Average temprature effect

```{r Temperature effect}
avg_temp <- Control_trees %>%
  group_by(Mean_temperature) %>%
  summarize(Mean_value = mean(log(Total), na.rm = TRUE))

model_Tmean <- lm(log(Total) ~ Day_of_year + Mean_temperature, 
                       data = Control_trees)
Table_S2 <- summary(model_Tmean)

# Table S2. Results of log-linear ANOVA analysis of average annual temperature effect on total monoterpene concentration
coef_df <- tidy(model_Tmean)
fit_df <- glance(model_Tmean)

coef_df <- coef_df[, c("term", "estimate", "std.error", "statistic", "p.value")]
names(coef_df) <- c("Term", "Estimate", "SE", "t", "p-value")
coef_df$`p-value` <- format.pval(
  coef_df$`p-value`,
  digits = 3,
  eps = 0.001
)
ft_coef <- flextable(coef_df)
ft_coef <- autofit(ft_coef)


fit_out <- data.frame(
  Statistic = c(
    "Residual SE",
    "R²",
    "Adjusted R²",
    "F statistic",
    "Model p-value"
  ),
  Value = c(
    round(fit_df$sigma, 3),
    round(fit_df$r.squared, 3),
    round(fit_df$adj.r.squared, 3),
    round(fit_df$statistic, 2),
    format.pval(fit_df$p.value, eps = 0.001)
  )
)

ft_fit <- autofit(flextable(fit_out))

doc <- body_add_break(doc)
doc <- body_add_par(
  doc,
  "Table S2. Results of log-linear ANOVA analysis of average annual temperature effect on total monoterpene concentration",
  style = "Normal"
)
doc <- body_add_par(doc, "Model: log(Total monoterpenes) ~ Day of year + Mean temperature", style = "Normal")
doc <- body_add_par(doc, "Model coefficients", style = "centered")
doc <- body_add_flextable(doc, ft_coef)
doc <- body_add_par(doc, "Model fit statistics", style = "centered")
doc <- body_add_flextable(doc, ft_fit)

Tmean_effect <- as.data.frame(effect("Mean_temperature", model_Tmean))

# Define parameters shown on the plot
slope <- round(exp(Table_S2$coefficients[grep("Mean_temperature", rownames(Table_S2$coefficients)), 1]), 3) # Slope coefficient
p_val <- round(Table_S2$coefficients[grep("Mean_temperature", rownames(Table_S2$coefficients)), 4],3) # p-value for slope
significance <- case_when(
  p_val < 0.001 ~ "p <0.001",
  p_val < 0.01  ~ "p <0.01",
  p_val < 0.05  ~ "p <0.05",
  TRUE          ~ ""
)

# Plot
Figure_3a <- ggplot(Tmean_effect, aes(x = Mean_temperature, y = exp(fit))) +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  geom_point(data = avg_temp, aes(x = Mean_temperature, y = exp(Mean_value),
                                  color = Climatedata$Mean_temperature, size = Climatedata$Elevation)) +  # Raw means in red
  scale_size_continuous(range = c(5, 12)) + # Adjust min and max sizes
  scale_color_gradient(low = "blue", high = "red") +
  annotate("text", x = min(Tmean_effect$Mean_temperature), y = min(exp(Tmean_effect$fit)), 
        label = as.expression(bquote("\nMultiplicative Coefficient =" ~ .(slope) ^ .(significance))),
           hjust = 0, vjust = -11, size = 6) +
  geom_errorbar(aes(x = avg_temp$Mean_temperature, ymin = exp(avg_temp$Mean_value)+exp(lower)-exp(fit), ymax = 
                      exp(avg_temp$Mean_value)+exp(upper)-exp(fit)), 
                width = 0.05, color = "darkblue") +  # Confidence intervals
    annotate("text", x = 1, y = 200, label = "a", size = 12) +

  labs(
    x = "Average Annual Temperature (\u00B0C)",
    y = expression('Monoterpene Concentration ' ~ (mu*g/g ~ FW)),
    color = "Temperature (\u00B0C)",
    size = "Elevation (M)"
  ) +
  theme_minimal() +
    theme(
    text = element_text(family = "Helvetica"),
    axis.text.x = element_text(size = 20, hjust = 1),
    axis.text.y = element_text(size = 20, hjust = 1),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 18),
    legend.text = element_text(size = 18),
    legend.title = element_text(size = 20),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
  )

rm(list = setdiff(ls(), c("Climatedata", "Monoterpenes", "Monoterpenes_Wide", "Control_trees", "doc", "Figure_3a", "Tmean_effect", "avg_temp")))
```

## Precipitation effect

```{r Precipitation effect}
avg_prep <- Control_trees %>%
  group_by(Precipitation) %>%
  summarize(Mean_value = mean(log(Total), na.rm = TRUE))
Control_trees$TreeID <- as.factor(Control_trees$TreeID)
model_PPT <- lm(log(Total) ~ Day_of_year + Precipitation, 
                       data = Control_trees)

Table_S3 <- summary(model_PPT)

# Table S3. Results of log-linear ANOVA analysis of annual precipitation effect on total monoterpene concentration
coef_df <- tidy(model_PPT)
fit_df <- glance(model_PPT)

coef_df <- coef_df[, c("term", "estimate", "std.error", "statistic", "p.value")]
names(coef_df) <- c("Term", "Estimate", "SE", "t", "p-value")
coef_df$`p-value` <- format.pval(
  coef_df$`p-value`,
  digits = 3,
  eps = 0.001
)
ft_coef <- flextable(coef_df)
ft_coef <- autofit(ft_coef)


fit_out <- data.frame(
  Statistic = c(
    "Residual SE",
    "R²",
    "Adjusted R²",
    "F statistic",
    "Model p-value"
  ),
  Value = c(
    round(fit_df$sigma, 3),
    round(fit_df$r.squared, 3),
    round(fit_df$adj.r.squared, 3),
    round(fit_df$statistic, 2),
    format.pval(fit_df$p.value, eps = 0.001)
  )
)

ft_fit <- autofit(flextable(fit_out))

doc <- body_add_break(doc)
doc <- body_add_par(
  doc,
  "Table S3. Results of log-linear ANOVA analysis of annual precipitation effect on total monoterpene concentration",
  style = "Normal"
)
doc <- body_add_par(doc, "Model: log(Total monoterpenes) ~ Day of year + Annual precipitation", style = "Normal")
doc <- body_add_par(doc, "Model coefficients", style = "centered")
doc <- body_add_flextable(doc, ft_coef)
doc <- body_add_par(doc, "Model fit statistics", style = "centered")
doc <- body_add_flextable(doc, ft_fit)

rm(list = setdiff(ls(), c("Climatedata", "Monoterpenes", "Monoterpenes_Wide", "Control_trees","doc", "Figure_3a", "Tmean_effect", "avg_temp")))
```

## VPD effect

```{r VPD effect}
avg_vpd <- Control_trees %>%
  group_by(Maximum_VPD) %>%
  summarize(Mean_value = mean(log(Total), na.rm = TRUE))

model_VPD <- lm(log(Total) ~ Day_of_year + Maximum_VPD, 
                       data = Control_trees)

Table_S4 <- summary(model_VPD)
# Table S4. Results of log-linear ANOVA analysis of annual maximum vapor pressure deficit effect on total monoterpene concentration
coef_df <- tidy(model_VPD)
fit_df <- glance(model_VPD)

coef_df <- coef_df[, c("term", "estimate", "std.error", "statistic", "p.value")]
names(coef_df) <- c("Term", "Estimate", "SE", "t", "p-value")
coef_df$`p-value` <- format.pval(
  coef_df$`p-value`,
  digits = 3,
  eps = 0.001
)
ft_coef <- flextable(coef_df)
ft_coef <- autofit(ft_coef)


fit_out <- data.frame(
  Statistic = c(
    "Residual SE",
    "R²",
    "Adjusted R²",
    "F statistic",
    "Model p-value"
  ),
  Value = c(
    round(fit_df$sigma, 3),
    round(fit_df$r.squared, 3),
    round(fit_df$adj.r.squared, 3),
    round(fit_df$statistic, 2),
    format.pval(fit_df$p.value, eps = 0.001)
  )
)

ft_fit <- autofit(flextable(fit_out))

doc <- body_add_break(doc)
doc <- body_add_par(
  doc,
  "Table S4. Results of log-linear ANOVA analysis of annual maximum vapor pressure deficit effect on total monoterpene concentration",
  style = "Normal"
)
doc <- body_add_par(doc, "Model: log(Total monoterpenes) ~ Day of year + Maximum VPD", style = "Normal")
doc <- body_add_par(doc, "Model coefficients", style = "centered")
doc <- body_add_flextable(doc, ft_coef)
doc <- body_add_par(doc, "Model fit statistics", style = "centered")
doc <- body_add_flextable(doc, ft_fit)


VPD_effect <- as.data.frame(effect("Maximum_VPD", model_VPD))

# Define parameters shown on the plot
slope <- round(exp(Table_S4$coefficients[grep("Maximum_VPD", rownames(Table_S4$coefficients)), 1]), 3) # Slope coefficient
p_val <- Table_S4$coefficients[grep("Maximum_VPD", rownames(Table_S4$coefficients)), 4] # p-value for slope
significance <- case_when(
  p_val < 0.001 ~ "p <0.001",
  p_val < 0.01  ~ "p <0.01",
  p_val < 0.05  ~ "p <0.05",
  TRUE          ~ ""
)
# Plot
Figure_3b <- ggplot(VPD_effect, aes(x = Maximum_VPD, y = exp(fit))) +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  geom_point(data = avg_vpd, aes(x = Maximum_VPD, y = exp(Mean_value),
                                  color = Climatedata$Mean_temperature, size = Climatedata$Elevation)) +  # Raw means in red
  scale_size_continuous(range = c(5, 12)) + # Adjust min and max sizes
  scale_color_gradient(low = "blue", high = "red") +
  annotate("text", x = min(VPD_effect$Maximum_VPD), y = min(exp(VPD_effect$fit)), 
        label = as.expression(bquote("\nMultiplicative Coefficient =" ~ .(slope) ^ .(significance))),
           hjust = 0, vjust = -11, size = 6) +
  geom_errorbar(aes(x = avg_vpd$Maximum_VPD, ymin = exp(avg_vpd$Mean_value)+exp(lower)-exp(fit), ymax =
                      exp(avg_vpd$Mean_value)+exp(upper)-exp(fit)), 
                width = 0.01, color = "darkblue") +  # Confidence intervals
    annotate("text", x = 0.7, y = 195, label = "b", size = 12) +

  labs(
    x = "Maximum vapor pressure deficit (kPa)",
    y = expression('Monoterpene Concentration ' ~ (mu*g/g ~ FW)),
    color = "Temperature (\u00B0C)",
    size = "Elevation (M)"
  ) +
  theme_minimal() +
  theme(
    text = element_text(family = "Helvetica"),
    axis.text.x = element_text(size = 20, hjust = 1),
    axis.text.y = element_text(size = 20, hjust = 1),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 18),
    legend.text = element_text(size = 18),
    legend.title = element_text(size = 20),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
  )

Figure_3 <- grid.arrange(Figure_3a, Figure_3b, ncol = 1)
ggsave("Figure_3.pdf", plot = Figure_3, width = 24, height = 30, units = "cm")

rm(list = setdiff(ls(), c("Climatedata", "Monoterpenes", "Monoterpenes_Wide", "Control_trees","doc")))
```


# 3.2. Survival as a function of constitutive and induced defense

Our analysis could not support this hypothesis. Defense induction was not affected by the environmental variables. Although colonized trees showed a high correlation with site elevation. 

## Extract baited trees in a new dataframe and add average control trees column as baseline and calculate fold change
```{r Calculate fold change}

# Separate baited trees from control in a new dataframe
Baited_trees <- Monoterpenes_Wide[Monoterpenes_Wide$Treatment == "Baited",]
# Drop COntrol level
Baited_trees$Status <- droplevels(Baited_trees$Status)
Baited_trees <- Baited_trees[, c(1:4,6:13,31)]

# Add Control trees' average by site and location as Baseline
Baited_trees <- merge(Baited_trees, 
                      setNames(aggregate(Total ~ Site + Location + Day_of_year, data = Control_trees, mean), 
                               c("Site", "Location", "Day_of_year", "Baseline")), 
                      by = c("Site", "Location", "Day_of_year"), all.x = TRUE)
Baited_trees <- Baited_trees %>%
  mutate(Site = factor(Site, levels = c("Grant", "Alma", "Empire", "Rabbit ears", "Cameron pass")))
# Calculate fold change
Baited_trees$Fold_change <- Baited_trees$Total/Baited_trees$Baseline
```


## Effect of environmental variables on induced response

To compare flight period induction I used day 170-190 concentrations.
To compare Post-flight period induction I used day 200-260 concentrations.

```{r Temperature effect on Induction}
# Flight period induction
Flight_induc <- lm(Fold_change ~ Day_of_year + Mean_temperature + Status, 
                       data = Baited_trees[Baited_trees$Day_of_year %in% c("170", "180", "190"),])
# Post-flight induction
Post_flight_induc <- lm(Fold_change ~ Day_of_year + Mean_temperature + Status, 
                       data = Baited_trees[Baited_trees$Day_of_year %in% c("200", "210", "225","240", "260"),])

# Table S5. Results of complete block ANOVA analysis of average annual temperature effect on induced fold change monoterpene concentration
coef_df <- tidy(Flight_induc)
fit_df <- glance(Flight_induc)

coef_df <- coef_df[, c("term", "estimate", "std.error", "statistic", "p.value")]
names(coef_df) <- c("Term", "Estimate", "SE", "t", "p-value")
coef_df$`p-value` <- format.pval(
  coef_df$`p-value`,
  digits = 3,
  eps = 0.001
)
ft_coef <- flextable(coef_df)
ft_coef <- autofit(ft_coef)


fit_out <- data.frame(
  Statistic = c(
    "Residual SE",
    "R²",
    "Adjusted R²",
    "F statistic",
    "Model p-value"
  ),
  Value = c(
    round(fit_df$sigma, 3),
    round(fit_df$r.squared, 3),
    round(fit_df$adj.r.squared, 3),
    round(fit_df$statistic, 2),
    format.pval(fit_df$p.value, eps = 0.001)
  )
)

ft_fit <- autofit(flextable(fit_out))

doc <- body_add_break(doc)
doc <- body_add_par(
  doc,
  "Table S5. Results of complete block ANOVA analysis of average annual temperature effect on induced fold change monoterpene concentration",
  style = "Normal"
)
doc <- body_add_par(doc,"")
doc <- body_add_par(doc, "Table S5-1: Flight period induced response", style = "Normal")
doc <- body_add_par(doc, "Model: Fold change ~ Day of year + Mean temperature + Status", style = "Normal")
doc <- body_add_par(doc, "Model coefficients", style = "centered")
doc <- body_add_flextable(doc, ft_coef)
doc <- body_add_par(doc, "Model fit statistics", style = "centered")
doc <- body_add_flextable(doc, ft_fit)


coef_df <- tidy(Post_flight_induc)
fit_df <- glance(Post_flight_induc)

coef_df <- coef_df[, c("term", "estimate", "std.error", "statistic", "p.value")]
names(coef_df) <- c("Term", "Estimate", "SE", "t", "p-value")
coef_df$`p-value` <- format.pval(
  coef_df$`p-value`,
  digits = 3,
  eps = 0.001
)
ft_coef <- flextable(coef_df)
ft_coef <- autofit(ft_coef)


fit_out <- data.frame(
  Statistic = c(
    "Residual SE",
    "R²",
    "Adjusted R²",
    "F statistic",
    "Model p-value"
  ),
  Value = c(
    round(fit_df$sigma, 3),
    round(fit_df$r.squared, 3),
    round(fit_df$adj.r.squared, 3),
    round(fit_df$statistic, 2),
    format.pval(fit_df$p.value, eps = 0.001)
  )
)

ft_fit <- autofit(flextable(fit_out))

doc <- body_add_par(doc,"")
doc <- body_add_par(doc, "Table S5-2: Post-flight period induced response", style = "Normal")
doc <- body_add_par(doc, "Model: Fold change ~ Day of year + Mean temperature + Status", style = "Normal")
doc <- body_add_par(doc, "Model coefficients", style = "centered")
doc <- body_add_flextable(doc, ft_coef)
doc <- body_add_par(doc, "Model fit statistics", style = "centered")
doc <- body_add_flextable(doc, ft_fit)

rm(list = setdiff(ls(), c("Climatedata", "Monoterpenes", "Monoterpenes_Wide", "Control_trees", "Baited_trees","doc")))
```

## Compare attack and mortality rate between the sites

Here I compare entry holes density between the sites along with the contingency table for beetle-killed and alive trees in each site.

```{r Attack rate and mortality}
Status_table <- Baited_trees %>%
  distinct(TreeID, .keep_all = TRUE) %>%  # Keep only one row per TreeID
  group_by(Site, Status) %>%
  summarise(Status_count = n(), .groups = "drop")  # Count occurrences per Site

# ANOVA for Entry holes density
model_Entryholes <- lm(Entryholes ~ Site, 
                       data = distinct(Baited_trees,TreeID, .keep_all = TRUE))

# Table S6. Results of ANOVA analysis on Entry hole density of trees in each site
coef_df <- tidy(model_Entryholes)
fit_df <- glance(model_Entryholes)

coef_df <- coef_df[, c("term", "estimate", "std.error", "statistic", "p.value")]
names(coef_df) <- c("Term", "Estimate", "SE", "t", "p-value")
coef_df$`p-value` <- format.pval(
  coef_df$`p-value`,
  digits = 3,
  eps = 0.001
)
ft_coef <- flextable(coef_df)
ft_coef <- autofit(ft_coef)


fit_out <- data.frame(
  Statistic = c(
    "Residual SE",
    "R²",
    "Adjusted R²",
    "F statistic",
    "Model p-value"
  ),
  Value = c(
    round(fit_df$sigma, 3),
    round(fit_df$r.squared, 3),
    round(fit_df$adj.r.squared, 3),
    round(fit_df$statistic, 2),
    format.pval(fit_df$p.value, eps = 0.001)
  )
)

ft_fit <- autofit(flextable(fit_out))

doc <- body_add_break(doc)
doc <- body_add_par(
  doc,
  "Table S6. Results of ANOVA analysis on Entry hole density of trees in each site",
  style = "Normal"
)
doc <- body_add_par(doc, "Model: Entryholes ~ Site", style = "Normal")
doc <- body_add_par(doc, "Model coefficients", style = "centered")
doc <- body_add_flextable(doc, ft_coef)
doc <- body_add_par(doc, "Model fit statistics", style = "centered")
doc <- body_add_flextable(doc, ft_fit)

# Plot Entry holes boxplot
Figure_4a <- ggplot(distinct(Baited_trees, TreeID, .keep_all = TRUE), 
             aes(x = Site, y = Entryholes)) +
  geom_boxplot(fill = "white", color = "black") +
  labs(x = element_blank(), 
       y = expression("Entry hole Density" ~ (m^-2))) +
  annotate("text", x = 0.8, y = 75, label = "a", size = 8) +
  coord_cartesian(clip = "off") +
theme_minimal() +
  theme(
    legend.position = "none", 
    axis.text.x = element_blank(), 
    panel.border = element_rect(color = "grey", fill = NA, linewidth = 1)
  )   # Turn off the legend

# Plot status bar chart
Figure_4b <- ggplot(Status_table, aes(x = Site, y = Status_count, fill = Status)) +
  geom_bar(
    stat = "identity",
    color = "black",     # bar border
    position = "stack"  # default; change if needed
  ) +
  scale_fill_grey(start = 0, end = 1) +
  scale_y_continuous(             # y-axis from 0 to 11
    breaks = 0:10                   # only integer values
  ) +
  labs(x = "Site", y = "Number of Trees") +
  annotate("text", x = 0.8, y = 9, label = "b", size = 8) +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.border = element_rect(color = "grey", fill = NA, linewidth = 1)
  )
# Arrange plots vertically
Figure_4 <- grid.arrange(Figure_4a, Figure_4b, ncol = 1)
ggsave("Figure_4.pdf", plot = Figure_4, width = 12, height = 18, units = "cm")

rm(list = setdiff(ls(), c("Climatedata", "Monoterpenes", "Monoterpenes_Wide", "Control_trees", "Baited_trees", "doc")))
```


## Concentration in trees under different status

```{r Concentration in trees}
model_status <- lm(Total ~ Day_of_year * Status, data = Monoterpenes_Wide)

# Get predictions with standard errors
predictions <- predict(model_status, newdata = Monoterpenes_Wide, se.fit = TRUE)

# Add predicted values and confidence intervals to the dataframe
Monoterpenes_Wide$Predicted <- predictions$fit
Monoterpenes_Wide$SE <- predictions$se.fit
Monoterpenes_Wide$CI_upper <- Monoterpenes_Wide$Predicted + 1.96 * Monoterpenes_Wide$SE
Monoterpenes_Wide$CI_lower <- Monoterpenes_Wide$Predicted - 1.96 * Monoterpenes_Wide$SE
Monoterpenes_Wide$Day_of_year <- as.character(Monoterpenes_Wide$Day_of_year)
Monoterpenes_Wide$Day_of_year <- as.numeric(Monoterpenes_Wide$Day_of_year)
position_jitter <- position_dodge(width = 3)
# Create the ggplot
Figure_5 <- ggplot(Monoterpenes_Wide, aes(x = Day_of_year, y = Predicted, linetype = Status, color = Status, group = Status)) +
  geom_line(linewidth = 1) +
  geom_point(position = position_jitter) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 4, linewidth = 0.9,position = position_jitter) +
  scale_linetype_manual(values = c("solid", "solid", "dashed", "dashed")) +
  scale_color_manual(values = c("black", "grey50", "black", "grey50")) +
  labs(
    x = "Day of year",
    y = expression('Total Monoterpene Concentration ' ~ (mu*g/g ~ FW))
  ) +
  theme_minimal() +
  theme(
    text = element_text(family = "Helvetica"),
    axis.text.x = element_text(size = 20, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 20, hjust = 1),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.title = element_text(size = 20),
    legend.text = element_text(size = 18),
    legend.position = c(0.8, 0.85),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
  )
# Save the plot
ggsave("Figure_5.pdf", plot = Figure_5, width = 35, height = 18, units = "cm")
Monoterpenes_Wide$Day_of_year <- as.factor(Monoterpenes_Wide$Day_of_year)

rm(list = setdiff(ls(), c("Climatedata", "Monoterpenes", "Monoterpenes_Wide", "Control_trees", "Baited_trees","doc")))
Monoterpenes_Wide <- Monoterpenes_Wide[,setdiff(names(Monoterpenes_Wide), c("Predicted", "SE", "CI_upper","CI_lower"))]
```

## Compare the status of baited trees

```{r Concentration in different status}
model_status <- lm(Fold_change ~ Day_of_year + Status, 
                       data = Baited_trees)

# Table S7. Result of Complete block ANOVA analysis of fold change concentration of monoterpenes in trees with different status
coef_df <- tidy(model_status)
fit_df <- glance(model_status)

coef_df <- coef_df[, c("term", "estimate", "std.error", "statistic", "p.value")]
names(coef_df) <- c("Term", "Estimate", "SE", "t", "p-value")
coef_df$`p-value` <- format.pval(
  coef_df$`p-value`,
  digits = 3,
  eps = 0.001
)
ft_coef <- flextable(coef_df)
ft_coef <- autofit(ft_coef)


fit_out <- data.frame(
  Statistic = c(
    "Residual SE",
    "R²",
    "Adjusted R²",
    "F statistic",
    "Model p-value"
  ),
  Value = c(
    round(fit_df$sigma, 3),
    round(fit_df$r.squared, 3),
    round(fit_df$adj.r.squared, 3),
    round(fit_df$statistic, 2),
    format.pval(fit_df$p.value, eps = 0.001)
  )
)

ft_fit <- autofit(flextable(fit_out))

doc <- body_add_break(doc)
doc <- body_add_par(
  doc,
  "Table S7. Result of Complete block ANOVA analysis of fold change concentration of monoterpenes in trees with different status",
  style = "Normal"
)

doc <- body_add_par(doc, "Model: Fold change ~ Day of year + Status", style = "Normal")
doc <- body_add_par(doc, "Model coefficients", style = "centered")
doc <- body_add_flextable(doc, ft_coef)
doc <- body_add_par(doc, "Model fit statistics", style = "centered")
doc <- body_add_flextable(doc, ft_fit)

# Add predicted values
Baited_trees$Predicted <- predict(model_status)

Status_effect <- as.data.frame(effect("Status", model_status))

# Tukey's HSD Test
tukey_status <- glht(model_status, linfct = mcp(Status = "Tukey"))
# Add letters to Status_effect dataframe
Status_effect$Significance <- cld(tukey_status)$mcletters$Letters


Figure_6 <- ggplot(Status_effect, aes(x = Status, y = fit)) +
  geom_point(size = 5) +  # Color based on Temperature
  geom_errorbar(aes(ymin = lower, ymax = upper), width = .10, color = "black", linewidth = 1) +  # Confidence intervals
  geom_text(aes(label = Significance, y = upper+.1), position = position_jitter(width = 0.0, height = 0.0), size = 10, color = "black") +  # Add significance letters
  labs(
    x = "Tree status",
    y = "FC - Total Monoterpene",
  ) +
  theme_minimal() +
    theme(
    text = element_text(family = "Helvetica"),
    axis.text.x = element_text(size = 20, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 20, hjust = 1),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.title = element_text(size = 20),
    legend.text = element_text(size = 18),
panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
  )

ggsave("Figure_6.pdf", plot = Figure_6, width = 25, height = 18, units = "cm")
rm(list = setdiff(ls(), c("Climatedata", "Monoterpenes", "Monoterpenes_Wide", "Control_trees", "Baited_trees","doc")))
```

## Logistic analysis between beetle-killed and living trees
For this analysis both colonized and live trees were pooled as living individuals.
Two different models were applied, one for during the flight period and one for after flight
```{r Monoterpene and survival}
# Create a new dataset
Baited_logistic_data <- Baited_trees[,c(1:5,15)]
Baited_logistic_data$Status <- as.character(Baited_logistic_data$Status)

# Change status to 0 and 1 for logistic analysis
Baited_logistic_data$Survival <- as.integer(Baited_logistic_data$Status != "Beetle-killed")


# Remove outliers larger than 5 fold change
Baited_logistic_data <- Baited_logistic_data[Baited_logistic_data$Fold_change <= 5, ]

# Divide dataset to seasons
Flight_data <- Baited_logistic_data[Baited_logistic_data$Day_of_year %in% c(160, 170, 180, 190),]
Postflight_data <- Baited_logistic_data[Baited_logistic_data$Day_of_year %in% c(200, 210, 225, 240, 260),]


# Combine data and model for constitutive and early season

Flight_logistic <- glm(Survival ~ Fold_change:Day_of_year, data = Flight_data, family = binomial)
Postflight_logistic <- glm(Survival ~ Fold_change:Day_of_year, data = Postflight_data, family = binomial)

Constitutive_odds <- round(exp(Flight_logistic$coefficients[[2]]),1)
Con_pval <- round(summary(Flight_logistic)$coefficients[2,4],2)

Day170 <- round(exp(Flight_logistic$coefficients[[3]]),1)
Day170pval <- round(summary(Flight_logistic)$coefficients[3,4],2)

Day180 <- round(exp(Flight_logistic$coefficients[[4]]),1)
Day180pval <- round(summary(Flight_logistic)$coefficients[4,4],2)

Day190 <- round(exp(Flight_logistic$coefficients[[5]]),1)
Day190pval <- round(summary(Flight_logistic)$coefficients[5,4],2)


# Table S8. Result of logistic regression showing monoterpene concentration fold change effect on tree survival
coef_df <- tidy(Flight_logistic)
fit_df <- glance(Flight_logistic)

coef_df <- coef_df[, c("term", "estimate", "std.error", "statistic", "p.value")]
names(coef_df) <- c("Term", "Estimate", "SE", "t", "p-value")
coef_df$`p-value` <- format.pval(
  coef_df$`p-value`,
  digits = 3,
  eps = 0.001
)
ft_coef <- flextable(coef_df)
ft_coef <- autofit(ft_coef)


fit_out <- data.frame(
  Statistic = c(
    "Null deviance",
    "Null DF",
    "Deviance",
    "Residual DF",
    "AIC"
  ),
  Value = c(
    round(fit_df$null.deviance, 3),
    round(fit_df$df.null, 3),
    round(fit_df$deviance, 3),
    round(fit_df$df.residual, 2),
    format.pval(fit_df$AIC, eps = 0.001)
  )
)

ft_fit <- autofit(flextable(fit_out))
doc <- body_add_break(doc)
doc <- body_add_par(
  doc,
  "Table S8. Result of logistic regression showing monoterpene concentration fold change effect on tree survival",
  style = "Normal"
)
doc <- body_add_par(doc,"")
doc <- body_add_par(doc, "Table S8-1: Flight period survival analysis", style = "Normal")
doc <- body_add_par(doc, "Model: Survival ~ Fold change × Day of year", style = "Normal")
doc <- body_add_par(doc, "Model coefficients", style = "centered")
doc <- body_add_flextable(doc, ft_coef)
doc <- body_add_par(doc, "Model fit statistics", style = "centered")
doc <- body_add_flextable(doc, ft_fit)


coef_df <- tidy(Postflight_logistic)
fit_df <- glance(Postflight_logistic)

coef_df <- coef_df[, c("term", "estimate", "std.error", "statistic", "p.value")]
names(coef_df) <- c("Term", "Estimate", "SE", "t", "p-value")
coef_df$`p-value` <- format.pval(
  coef_df$`p-value`,
  digits = 3,
  eps = 0.001
)
ft_coef <- flextable(coef_df)
ft_coef <- autofit(ft_coef)


fit_out <- data.frame(
  Statistic = c(
    "Null deviance",
    "Null DF",
    "Deviance",
    "Residual DF",
    "AIC"
  ),
  Value = c(
    round(fit_df$null.deviance, 3),
    round(fit_df$df.null, 3),
    round(fit_df$deviance, 3),
    round(fit_df$df.residual, 2),
    format.pval(fit_df$AIC, eps = 0.001)
  )
)

ft_fit <- autofit(flextable(fit_out))

doc <- body_add_par(doc,"")
doc <- body_add_par(doc, "Table S8-2: Post-flight period survival analysis", style = "Normal")
doc <- body_add_par(doc, "Model: Survival ~ Fold change × Day of year", style = "Normal")
doc <- body_add_par(doc, "Model coefficients", style = "centered")
doc <- body_add_flextable(doc, ft_coef)
doc <- body_add_par(doc, "Model fit statistics", style = "centered")
doc <- body_add_flextable(doc, ft_fit)
print(doc, "Supplementary.docx")

rm(list = setdiff(ls(), c("Baited_logistic_data", "Monoterpenes_Wide", "Control_trees", "Baited_trees","doc","Con_pval",
                          "Constitutive_odds", "Day170", "Day170pval", "Day180", "Day180pval", "Day190", "Day190pval")))
```

## Graph of the seasonal survival analysis

```{r Final graph of survival}
model_status_int <- lm(Fold_change ~ Day_of_year * Survival, 
                       data = Baited_logistic_data)

# Add predicted values and confidence intervals to the dataframe
Baited_logistic_data$Predicted <- predict(model_status_int, newdata = Baited_logistic_data)
Baited_logistic_data$Survival[Baited_logistic_data$Survival == "0"] <- "Killed"
Baited_logistic_data$Survival[Baited_logistic_data$Survival == "1"] <- "Survived"
# Create the ggplot choose colors with this code: grep("yellow", colors(), value = TRUE)
Figure_7 <- ggplot(Baited_logistic_data, aes(x = factor(Day_of_year), y = Fold_change, fill = Survival)) +

    # Add shaded regions
  geom_rect(aes(xmin = 0.5, xmax = 4.5, ymin = 0, ymax = 5), fill = "white", alpha = 0.008, inherit.aes = FALSE,) +
  geom_rect(aes(xmin = 4.5, xmax = 9.5, ymin = 0, ymax = 5), fill = "gray", alpha = 0.008, inherit.aes = FALSE) +
  geom_boxplot() +
  scale_fill_manual(values = c("gray", "white")) +

  # Add attack date
  geom_vline(xintercept = 1.7, linetype = "dashed", color = "black", linewidth = 1) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "black", size = 1) +
  # Add text labels for shaded regions
  annotate("text", x = 2.3, y = 4, label = "Flight period", size = 6) +
  annotate("text", x = 7, y = 4, label = "Post-flight period", size = 6) +
  
  # Add odds values
  annotate("text", x = 1, y = 3.5, 
    label = as.expression(bquote("Odds =" ~ .(Constitutive_odds)), size = 6)) +
  annotate("text", x = 1, y = 3.3, 
    label = as.expression(bquote("P =" ~ .(Con_pval)), size = 6)) +
  annotate("text", x = 2, y = 3.5, 
    label = as.expression(bquote("Odds =" ~ .(Day170)), size = 6)) +
  annotate("text", x = 2, y = 3.3, 
    label = as.expression(bquote("P =" ~ .(Day170pval)), size = 6)) +
  annotate("text", x = 3, y = 3.5, 
    label = as.expression(bquote("Odds =" ~ .(Day180)), size = 6)) +
  annotate("text", x = 3, y = 3.3, 
    label = as.expression(bquote("P =" ~ .(Day180pval)), size = 6)) +
  annotate("text", x = 4, y = 3.5, 
    label = as.expression(bquote("Odds* =" ~ .(Day190)), size = 6)) +
  annotate("text", x = 4, y = 3.3, 
    label = as.expression(bquote("P =" ~ .(Day190pval)), size = 6)) +
  
  annotate("text", x = 5, y = 3.5, label = "ns", size = 5) +
  annotate("text", x = 6, y = 3.5, label = "ns", size = 5) +
  annotate("text", x = 7, y = 3.5, label = "ns", size = 5) +
  annotate("text", x = 8, y = 3.5, label = "ns", size = 5) +
  annotate("text", x = 9, y = 3.5, label = "ns", size = 5) +

  # Define labels
  labs(
    x = "Day of year",
    y = "Fold Change in Total Monoterpene") +
  theme_minimal() +
  ylim(0, 5) +
  theme(
    text = element_text(family = "Helvetica"),
    axis.text.x = element_text(size = 20, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 20, hjust = 1),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.title = element_text(size = 20),
    legend.text = element_text(size = 18),
    legend.position = c(0.9, 0.85),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    panel.grid.major = element_line(color = "gray70", linewidth = 1),  # Major grid lines
    panel.grid.minor = element_line(color = "gray90", linewidth = 1),  # Major grid lines

  )

ggsave("Figure_7.pdf", plot = Figure_7, width = 40, height = 20, units = "cm")
rm(list = setdiff(ls(), c("Baited_logistic_data", "Monoterpenes_Wide", "Control_trees", "Baited_trees","doc")))
```

# 3.3 Interaction between time and concentration

```{r Interaction between speed and magnitude}
Baited_interaction <- Baited_logistic_data[,-c(5, 8)]

Baited_int_wide <- Baited_interaction %>% 
  pivot_wider(
    names_from = Day_of_year,
    values_from = Fold_change
  )
  
Baited_int_wide$Magnitude <- rowMeans(Baited_int_wide[,5:13], na.rm = TRUE)
#Baited_int_wide$MaxFold_change <- do.call(pmax, c(Baited_int_wide[, 5:13], na.rm = TRUE))

Baited_int_wide$Speed <- apply(as.data.frame(Baited_int_wide[,5:13]>1), 1, function(x) {
  idx <- which(x)
  if (length(idx) == 0) NA else names(as.data.frame(Baited_int_wide[,5:13]>1))[idx[1]]
})
Baited_int_wide$Survival[Baited_int_wide$Survival == "Survived"] <- "1"
Baited_int_wide$Survival[Baited_int_wide$Survival == "Killed"] <- "0"
Baited_int_wide$Survival <- as.numeric(Baited_int_wide$Survival)
Baited_int_wide$Speed <- as.factor(Baited_int_wide$Speed)

Effect_logistic <- glm(Survival ~ Speed*Magnitude, data = Baited_int_wide, family = binomial)

# Table S9. Result of logistic regression test of speed and magnitude of monoterpene induction on survival
coef_df <- tidy(Effect_logistic)
fit_df <- glance(Effect_logistic)

coef_df <- coef_df[, c("term", "estimate", "std.error", "statistic", "p.value")]
names(coef_df) <- c("Term", "Estimate", "SE", "t", "p-value")
coef_df$`p-value` <- format.pval(
  coef_df$`p-value`,
  digits = 3,
  eps = 0.001
)
ft_coef <- flextable(coef_df)
ft_coef <- autofit(ft_coef)


fit_out <- data.frame(
  Statistic = c(
    "Null deviance",
    "Null DF",
    "Deviance",
    "Residual DF",
    "AIC"
  ),
  Value = c(
    round(fit_df$null.deviance, 3),
    round(fit_df$df.null, 3),
    round(fit_df$deviance, 3),
    round(fit_df$df.residual, 2),
    format.pval(fit_df$AIC, eps = 0.001)
  )
)

ft_fit <- autofit(flextable(fit_out))

doc <- body_add_break(doc)
doc <- body_add_par(
  doc,
  "Table S9. Result of logistic regression test of speed and magnitude of monoterpene induction on survival",
  style = "Normal"
)
doc <- body_add_par(doc,"")
doc <- body_add_par(doc, "Model: Survival ~ Speed × Magnitude", style = "Normal")
doc <- body_add_par(doc, "Model coefficients", style = "centered")
doc <- body_add_flextable(doc, ft_coef)
doc <- body_add_par(doc, "Model fit statistics", style = "centered")
doc <- body_add_flextable(doc, ft_fit)
doc <- body_add_par(doc, "* NA in some rows is the result of having only one replicates in those factor levels", style = "Normal")
print(doc, "Supplementary.docx")
```

